---
title: "`pwr` cheat sheet"
author: "Clay Ford"
output: pdf_document
---

The `pwr` package provides basic functions for power and sameple size analysis as described in the book _Statistical power analysis for the behavioral sciences_ (2nd ed.) by Jacob Cohen. 

```{r}
library(pwr)
```

### conventional effect sizes for certain tests

Test                        | `small` | `medium` | `large`
----------------------------|-------|--------|-------
tests for proportions (`p`) | 0.2   | 0.5    | 0.8
tests for means (`t`)       | 0.2   | 0.5    | 0.8
chi-square tests (`chisq`)| 0.1   | 0.3    | 0.5
correlation test (`r`)      | 0.1   | 0.3    | 0.5 
anova (`anov`)                 | 0.1   | 0.25   | 0.4 
general linear model (`f2`) | 0.02  | 0.15   | 0.35 

To get these in R: 
`cohen.ES(test = c("p", "t", "r", "anov", "chisq", "f2"), size = c("small", "medium", "large"))`

Example:
```{r}
cohen.ES(test = "anov", size = "small")
```


### one-sample test for proportions  

Say we think people place name tags on 
the left side of their chest 65% percent of the time versus random chance
(50%). What sample size do we need to show this assuming a significance level
(Type I error) of 0.05 and a desired power of 0.80?

```{r eval=FALSE}
pwr.p.test(h = ES.h(p1 = 0.65, p2 = 0.50), sig.level = 0.05, 
           power = 0.80, alternative = "greater")
```

Say we think people place name tags on the left 75% percent of the time instead 
of 50%. What is the power of our test if we survey 30 people provided
we accept a significance level (Type I error) of 0.05? 
```{r eval=FALSE}
pwr.p.test(h = ES.h(p1 = 0.75, p2 = 0.50), n = 30, 
           sig.level = 0.05, alternative = "greater")
```

Say we think people placing name tags on the left side of their chest is not governed by random chance (50%). What sample size do we need to detect a "small" effect of 0.2 assuming a significance level of 0.05 and a desired power of 0.80?

```{r eval=FALSE}
pwr.p.test(h = 0.2, sig.level = 0.05, power = 0.8, alternative = "two.sided")
```

### two-sample test for proportions

I randomly sample male and female UVa undergrad students
and ask them if they consume alcohol at least once a week. My null hypothesis
is no difference in the proportion that answer yes. My alternative hypothesis
is that there is a difference. (two-sided; one gender has higher proportion, I
don't know which.) I'd like to detect a difference as small as 5%. How many
students do I need to sample in each group if we want 80% power and 5% chance
of Type 1 error?

```{r eval=FALSE}
# 55% vs. 50%
pwr.2p.test(h = ES.h(p1 = 0.55, p2 = 0.50), sig.level = 0.05, power = .80)
# 35% vs. 30%
pwr.2p.test(h = ES.h(p1 = 0.35, p2 = 0.30), sig.level = 0.05, power = .80)
# using "small" conventional effect size 
pwr.2p.test(h = 0.2, sig.level = 0.05, power = .80)
```


### two-sample test for proportions, unequal sample sizes

I randomly sample male and female UVa undergrad students
and ask them if they consume alcohol at least once a week. My null hypothesis
is no difference in the proportion that answer yes. My alternative hypothesis
is that there is a difference. (two-sided; one gender has higher proportion, I
don't know which.) We were able to survey 543 males and 675 females. What's the power of our
test if we want to be able to detect a "small" effect (h = 0.2)?
```{r eval=FALSE}
pwr.2p2n.test(h = 0.2, n1 = 543, n2 = 675, sig.level = 0.05)
```


### one-sample and two-sample t-tests for means

The effect size for two-sample t-tests:

$$ d = \frac{m_1 - m_2}{\sigma}$$

The effect size for one-sample t-tests:

$$ d = \frac{m_1 - null}{\sigma}$$

The effect size for paired t-tests (with and without correlation):

$$ d = \frac{m_1 - m_2}{\sigma \sqrt{2}}$$
$$ d = \frac{m_1 - m_2}{\sigma \sqrt{2(1 - r)}}$$


I'm interested to know if there is a difference in the mean price of what male
and female students pay at the library coffee shop. Let's say I randomly 
observe 30 male and 30 female students check out from the coffee shop and note
their total purchase price. How powerful is this experiment if I want to
detect a "small" effect in either direction?

```{r eval=FALSE}
# two-sample test
pwr.t.test(n = 30, d = 0.2, sig.level = 0.05) # n is per group

```

How many do I need to observe for a test with 80% power?
```{r eval=FALSE}
pwr.t.test(d = 0.2, power = 0.80, sig.level = 0.05)
# alternative: greater than 0 (positive effect)
pwr.t.test(d = 0.2, power = 0.80, sig.level = 0.05, alternative = "greater")
# alternative: less than 0 (negative effect)
pwr.t.test(d = -0.2, power = 0.80, sig.level = 0.05, alternative = "less")
```

Let's say we want to be able to detect a difference of at least 75 cents in 
the mean purchase price. How can we convert that to an effect size? First we 
need to make a guess at the population standard deviation. If we have 
absolutely no idea, one rule of thumb is to take the difference between the 
maximum and minimum values and divide by 4 (or 6). Let's say max is 10 and min
is 1. So our guess at a standard deviation is 9/4 = 2.25. 

```{r eval=FALSE}
# one-sample test
d <- 0.75/2.25 # 0.333
pwr.t.test(d = d, power = 0.80, sig.level = 0.05, type = "one.sample")

```

24 high school boys are put on a ultraheavy rope-jumping program. 
Does this increase their 40-yard dash time? We'll measure their 40 time before
the program and after. We'll use a paired t-test to see if the difference in 
times is greater than 0. Assume the standard deviation of the differences will
be about 0.25. How powerful is the test to detect a difference of about positive 0.08
with 0.05 significance?

```{r eval=FALSE}
# paired t-test
pwr.t.test(n = 24, d = 0.08 / (0.25 * sqrt(2)), 
           type = "paired", alternative = "greater")

# or 
power.t.test(n = 24, delta = 0.08, sd = 0.25 * sqrt(2), 
             type = "paired", alternative = "one.sided")

```



### two sample t test for means, unequal sample sizes

Find power for a t test with 28 in one group and 35 in the other group and a
medium effect size. (sig.level defaults to 0.05.)
```{r eval=FALSE}
pwr.t2n.test(n1 = 28, n2 = 35, d = 0.5)

```


Find n1 sample size when other group has 35, desired power is 0.80, effect 
size is 0.5 and significance level is 0.05:
```{r eval=FALSE}
pwr.t2n.test(n2 = 35, d = 0.5, power = 0.8)
```

### chi-squared tests 

(From Cohen, example 7.1) A market researcher is seeking to determine
preference among 4 package designs. He arranges to have a panel of 100 
consumers. He wants to perform a chi-square goodness of fit test against the 
null of equal preference (25% for each design) with a significance level of 
0.05. What's the power of the test if 3/8 of the population actually prefers 
one of the designs and the remaining 5/8 are split over the other 3 designs?

```{r eval=FALSE}
# Goodness of Fit
# We need to create vectors of null and alternative proportions:
P0 <- rep(0.25, 4)
P1 <- c(3/8, rep((5/8)/3, 3))

# To calculate power, specify effect size, N, and degrees of freedom (4-1).
pwr.chisq.test(w=ES.w1(P0,P1), N=100, df=(4-1), sig.level=0.05)

# How many subjects do we need to achieve 80% power?
pwr.chisq.test(w=ES.w1(P0,P1), df=(4-1), power=0.8, sig.level = 0.05)

```

I want to see if there's an association between gender and flossing
teeth among UVa students. I randomly sample 100 students (male and female) and
ask whether or not they floss daily. I want to carry out a chi-square test of 
association to determine if there's an association between these two 
variables. As usual I set my significance level to 0.05. To determine effect 
size I need to propose an alternative hypothesis, which in this case is a 
table of proportions.

```{r eval=FALSE}
# test of association
# using hypothesized proportions
prob <- matrix(c(0.10,0.20,0.40,0.30), ncol=2, 
               dimnames = list(c("M","F"),c("Floss","No Floss")))
pwr.chisq.test(w = ES.w2(prob), N = 100, df = 1, sig.level = 0.05)

# How many students should I survey if I wish to achieve 90% power?
pwr.chisq.test(w = ES.w2(prob), power = 0.9, df = 1, sig.level = 0.05)

# using "small" conventional effect
pwr.chisq.test(w = 0.1, N = 100, df = 1, sig.level = 0.05)
pwr.chisq.test(w = 0.1, power = 0.9, df = 1, sig.level = 0.05)
```

### correlation test

Let's say I'm a web developer and I want to conduct an experiment with one of 
my sites. I want to randomly select a group of people, ranging in age from 18 
- 65, and time them how long it takes them to complete a task, say locate some
piece of information. I suspect there may be a "small" positive linear
relationship between time it takes to complete the task and age. How many
subjects do I need to detect this relationship with 80% power and the usual
0.05 significance level?

```{r eval=FALSE}
# "small" effect is 0.1
pwr.r.test(r = 0.1, sig.level = 0.05, power = 0.8, alternative = "greater")
# detect small effect in either direction 
pwr.r.test(r = 0.1, sig.level = 0.05, power = 0.8, alternative = "two.sided")

```

### balanced one-way analysis of variance tests

Let's say I'm a web developer and I'm interested in 3 web site 
designs for a client. I'd like to know which design(s) help users find 
information fastest, or which design requires the most time. I design an 
experiment where I have 3 groups of randomly selected people use one of the 
designs to find some piece of information and I record how long it takes. (All
groups look for the same information.) How many people do I need in each group
to detect a "medium" effect if I desire power
and significance levels of 0.8 and 0.05?

```{r eval=FALSE}
pwr.anova.test(k = 3, f = 0.25, sig.level = 0.05, power = 0.8)
```

What's the power of my test to detect a medium effect if I get 50 for each group?

```{r eval=FALSE}
pwr.anova.test(k = 3, f = 0.25, sig.level = 0.05, n = 50)
```

How many people do I need in each group
if I believe two of the designs will take 30 seconds and one will take 35
seconds? Assume population standard deviation is 10 and that I desire power
and significance levels of 0.8 and 0.05. Using `power.anova.test`:

```{r eval=FALSE}
power.anova.test(groups = 3, between.var = var(c(30, 30, 35)), 
                 within.var = 10^2, power = 0.8)

```


### test for the general linear model

Let's say I'm a web developer and I want to conduct an experiment with one of 
my sites. I want to randomly select a group of people, ranging in age from 18 
- 65, and time them how long it takes them to complete a task, say locate some
piece of information. I know there will be variability in the observed times. 
I think age, gender and years of education may explain this variability. How 
powerful is my experiment if I recruit 40 subjects and I want to be able to 
detect at least 30% explained variance ($R^{2} = .30$) with a 0.05 significance 
level?

Three predictors, so u = 3. 40 subjects, so v = 40 - 3 - 1. $R^{2} = .30$, so effect size is $f2 = 0.3/(1 - 0.3)$

```{r eval=FALSE}
pwr.f2.test(u = 3, v = 40 - 3 - 1, f2 = 0.3/(1 - 0.3), sig.level = 0.05)

```


How many subjects do I need if I want to be able to detect at least 30%
explained variance ($R^{2} = .30$) with 80% power and the usual 0.05 significance
level? We have to find v and than derive n.

```{r eval=FALSE}
pwr.f2.test(u = 3, f2 = 0.3/(1 - 0.3), sig.level = 0.05, power = 0.8)
# n = round(v) + u + 1
```

### References


Cohen, J. (1988). _Statistical Power Analysis for the Behavioral Sciences (2nd ed.)_. LEA.  

Dalgaard, P. (2002). _Introductory Statistics with R_. Springer.   

Hogg, R and Tanis, E. (2006). _Probability and Statistical Inference (7th ed.)_. Pearson.  

Ryan, T. (2013). _Sample Size Determination and Power_. Wiley.   
